# Programming Assigent 1. Reproducible Research

Author: Ruben Sanchez

### Loading and transforming the dataset

We'll first load the csv file

```{r, echo=TRUE, results='asis'}
fitdata <- read.csv("activity.csv")
```

And now we transform the date field and complete the times to four positions

```{r, echo=TRUE, results='asis'}
fitdata$date <- as.Date(fitdata$date)
fitdata$interval <- sprintf("%04d", fitdata$interval)
str(fitdata)
```

### Steps by day

We are now going to sum the total steps for each day and see it in a histogram

```{r, echo=TRUE}
fitdata_group <- tapply(fitdata$steps, fitdata$date, sum)
library(ggplot2)
qplot(fitdata_group, xlab = "Steps by day")
```

Let's calculate the mean and median of the steps per day

```{r, echo=TRUE}
mean(fitdata_group, na.rm = TRUE)
median(fitdata_group, na.rm = TRUE)
```

### Average daily activity pattern

Next, we are going to calculate the mean number of steps in each 5-minute interval,
and we'll show it with a time series plot and looking for the maximum.

```{r, echo=TRUE}
fitdata_time <- tapply(fitdata$steps, fitdata$interval, mean, na.rm =TRUE)
times <- unique(fitdata$interval)
fitdata_time <- cbind(fitdata_time, times)
plot(fitdata_time[,2], fitdata_time[,1], type = "l", xlab = "interval", ylab = "steps")
interval_max <- fitdata_time[which.max(fitdata_time[,1]),2]
print(paste("The 5-minute interval with the maximum steps is", interval_max))
```

### Imputing missing values

Let's start by calculating the number of NAs we have

```{r, echo=TRUE}
total_na <- sum(is.na(fitdata$steps))
total_na
```

To fill the not available values we are going to use the mean steps in each interval

```{r, echo=TRUE}
fitdata2 <- fitdata
fitdata2$steps[is.na(fitdata2$steps)] <- tapply(fitdata$steps, fitdata$interval, mean, na.rm =TRUE)
fitdata_group2 <- tapply(fitdata2$steps, fitdata2$date, sum)
qplot(fitdata_group2, xlab = "Steps by day")
mean(fitdata_group2)
median(fitdata_group2)
```

We can see there is not a great difference between these results and the ones we obtained at the beginning of the analysis

### Differences in weekdays

We add a new fied for the type of day and fill it with weekday or weekend

```{r, echo=TRUE}
library(dplyr)
fitdata2$day <- weekdays(fitdata2$date)
fitdata2$day[fitdata2$day == "Saturday" | fitdata2$day == "Sunday"] <- "weekend"
fitdata2$day[fitdata2$day == "Monday" | fitdata2$day == "Tuesday"| fitdata2$day == "Wednesday"| fitdata2$day == "Thursday"| fitdata2$day == "Friday"] <- "weekday"
fitdataday <- filter(fitdata2, day == "weekday")
fitdataend <- filter(fitdata2, day == "weekend")
fitdata_timeday <- tapply(as.numeric(fitdataday$steps), fitdataday$interval, mean, na.rm =TRUE)
fitdata_timeend <- tapply(as.numeric(fitdataend$steps), fitdataend$interval, mean, na.rm =TRUE)
times2 <- unique(fitdata2$interval)
fitdata_timeday <- cbind(fitdata_timeday, times2, "weekday")
fitdata_timeend <- cbind(fitdata_timeend, times2, "weekend")
fitdata_time2 <- rbind(fitdata_timeday, fitdata_timeend)
library(lattice)
xyplot(as.numeric(fitdata_time2[,1])~as.numeric(fitdata_time2[,2]) | fitdata_time2[,3], layout = c(1,2), type = "l", lty = 1, xlab = "Intervals", ylab = "Steps")
```

We can see several differences between weekdays and weekends. People start to walk earlier at weekdays, but later the number of steps decreases a lot.